My Publication Title --- Single Author

### 2025-07-17 paper/quarto_aaai/paper.tex
removed figure and synthetic example from experiments

Today's prominence of artificial intelligence (AI) has largely been
driven by \textbf{representation learning}: instead of relying on
features and rules hand-crafted by humans, modern machine learning (ML)
models are tasked with learning representations directly from the data,
guided by narrow objectives such as predictive accuracy
\citep{goodfellow2016deep}. Advances in computing have made it possible
to provide these models with ever-growing degrees of freedom to achieve
this task, which often allows them to outperform traditionally
parsimonious models. Unfortunately, in doing so, models learn
increasingly complex, sensitive representations that humans can no
longer easily interpret.

The trend towards complexity for the sake of performance has come under
scrutiny in recent years. At the very cusp of the deep learning (DL)
revolution, \citet{szegedy2013intriguing} showed that artificial neural
networks (ANN) are susceptible to adversarial examples (AEs): perturbed
versions of data instances that yield vastly different model predictions
despite being semantically indistinguishable from their factual
counterparts. Some partial mitigation strategies have been
proposed---most notably \textbf{adversarial training}
\citep{goodfellow2014explaining}---but truly robust deep learning
remains unattainable even for models that are considered ``shallow'' by
today's standards \citep{kolter2023keynote}.

Part of the problem is that the high degrees of freedom---high number of
parameters estimated from data---provide room for many solutions that
are locally optimal with respect to narrow objectives
\citep{wilson2020case}. As one example, research on the ``lottery ticket
hypothesis'' suggests that modern neural networks can be pruned by up to
90\% without losing predictive performance \citep{frankle2018lottery}.
Thus, looking at the predictive performance alone, found solutions may
seem to provide compelling explanations for the data, when in fact they
are based on purely associative and semantically meaningless patterns.
This poses two related challenges. Firstly, there is no dependable way
to verify if learned representations correspond to meaningful, plausible
explanations. Secondly, even if we resolve this challenge, it remains
undecided how to ensure that machine learning models can \emph{only}
learn valuable explanations.

The first challenge has attracted an abundance of work on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations, aiming to mitigate scenarios in which
practitioners deploy opaque models and have to blindly rely on their
predictions. On many occasions, this has happened in practice, causing
harms to people who were adversely and unfairly affected by automated
decision-making (ADM) systems involving opaque models; see, e.g.,
\citet{oneil2016weapons}. Effective XAI tools can also aid in monitoring
models and providing recourse, empowering people to turn negative
outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
``loan application accepted''). In line with this, our work builds upon
\textbf{counterfactual explanations} (CE) proposed by
\citet{wachter2017counterfactual}; CEs prescribe minimal changes for
factual inputs that, if implemented, would prompt some fitted model to
produce an alternative, more desirable output.

To our surprise, the second challenge has not yet attracted major
research interest. In particular, there has been no concerted effort
towards improving the degree to which learned representations promote
explanations that are both \textbf{interpretable} to and deemed
\textbf{plausible} by humans. Instead, the typical choice has been to
improve the ability of XAI tools to identify the subset of explanations
that are plausible and valid for any given model, independent of whether
these explanations are compatible with the learned representations
\citep{altmeyer2024faithful}. Fortunately, recent findings indicate that
improved ``explanatory capacity'' of a model can arise as a consequence
of regularization techniques aimed at other training objectives such as
generative capacity, generalization, or robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
Our contribution consolidates these findings within a unified framework.

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explanatory
capacity of models that, in high-level terms, we define as the extent to
which valid counterfactual explanations derived for a model can be
deemed plausible with respect to the underlying data and global
actionability constraints (we refine this notion in Def.
\ref{def-explainability}). To the best of our knowledge, Counterfactual
Training represents the first attempt to achieve more explainable models
by employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we empirically demonstrate that CT substantially
improves explainability and positively contributes to the robustness of
trained models without sacrificing predictive performance. Finally, in
Section~\ref{sec-discussion}, we discuss open challenges and conclude
that CT is a promising approach towards making opaque models more
trustworthy.

@@

To make the desiderata for our framework more concrete, we follow
\citet{augustin2020adversarial} in tying explainability to the quality
of CEs that can be generated for a given model. The authors show that
CEs (understood as minimal input perturbations that yield some desired
model prediction) tend to be more meaningful if the underlying model is
more robust to adversarial examples. We can make intuitive sense of this
finding if we look at adversarial training (AT) through the lens of
representation learning with high degrees of freedom. As argued before,
learned representations may be sensitive to producing implausible
explanations and mispredicting for worst-case counterfactuals (i.e.,
AEs). Thus, by inducing models to ``unlearn'' susceptiblity to such
examples, adversarial training can effectively remove implausible
explanations from the solution space.

\subsection{Adversarial Examples are
Counterfactuals}\label{adversarial-examples-are-counterfactuals}

The interpretation of the link between explainability through
counterfactuals on the one side, and robustness to adversarial examples
on the other is backed by empirical evidence.
\citet{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\citet{abbasnejad2020counterfactual} argue that counterfactuals
represent potentially useful training data in machine learning,
especially in supervised settings where inputs may be reasonably mapped
to multiple outputs. They, too, show that augmenting the training data
of (image) classifiers can improve generalization performance. Finally,
\citet{teney2020learning} argue that counterfactual pairs tend to exist
in training data. Hence, their approach aims to identify similar input
samples with different annotations and ensure that the gradient of the
classifier aligns with the vector between such pairs of counterfactual
inputs using a cosine distance loss function.

@@

\subsection{Aligning Representations with
Explanations}\label{aligning-representations-with-explanations}

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a
closely related work, \citet{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. They propose a way to generate counterfactuals that are
maximally faithful to the model in that they are consistent with what
the model has learned about the underlying data. Formally, they rely on
tools from energy-based modelling \citep{teh2003energy} to minimize the
divergence between the distribution of counterfactuals and the
conditional posterior over inputs learned by the model. Their
counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
and only if the underlying model has learned representations that align
with them. The authors find that both deep ensembles
\citep{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\citep{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are particularly effective
when models are underspecified by the available data
\citep{wilson2020case}. Averaging across solutions mitigates the
aforementioned risk of overrelying on a single locally optimal
representation that corresponds to semantically meaningless explanations
for the data. Likewise, previous work of \citet{schut2021generating}
found that generating plausible (``interpretable'') CEs is almost
trivial for deep ensembles that have undergone adversarial training. The
case for JEMs is even clearer: they optimize a hybrid objective that
induces both high predictive performance and strong generative capacity
\citep{grathwohl2020your}, which bears resemblance to the idea of
aligning models with plausible explanations and has inspired our CT
objective.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability: (a) conventional training, all mutable; (b) CT, all
mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
immutable. The linear decision boundary is shown in green along with
training data colored according to ground-truth labels: \(y^-=1\) (blue)
and \(y^+=2\) (orange). Stars indicate counterfactuals in the target
class.}

\subsubsection{Synthetic Example: Prediction of Credit Card
Defaults.}\label{synthetic-example-prediction-of-credit-card-defaults.}

Figure~\ref{fig-poc} presents results for a linear classifier fitted to
\emph{LS} that complies with the data assumptions in Proposition
\ref{prp-mtblty}. The four panels show the outcomes for different
training procedures: in panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we have applied CT. For
illustrative purposes, suppose the first feature represents \emph{debt}
(mutable) and the second feature represents \emph{age} (immutable) of
loan applicants seeking counterfactual explanations for moving to the
target class: loan provided (orange).

In all four cases, it is possible to generate valid counterfactuals
(stars) for unsuccessful applicants (blue). They cross the decision
boundary (green) into the target class, but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly discernible cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. This
difference in outcomes is quantified for the non-linear MLP in the first
two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
implausibility: it is substantial and statistically significant for
\emph{LS} across both metrics, the distance-based \(\text{IP}\) (29\%)
and divergence-based \(\text{IP}^*\) (55\%).

In panel (c) of Figure~\ref{fig-poc}, the CEs involve substantial
reductions in \emph{debt} for younger applicants. By comparison,
counterfactual paths are shorter on average in panel (d) where we have
protected the immutable \emph{age} as described in
Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
to \emph{age}, recommendations with respect to \emph{debt} are much more
homogenous and do not unfairly punish younger individuals. These CEs are
also plausible with respect to the mutable feature, despite requiring
smaller debt reductions on average, resulting in smaller costs to
individuals. This result is quantified for the non-linear case in column
three of Table~\ref{tbl-main}, which shows the \(\%\)-reduction in costs
averaged across valid counterfactuals. Once again, the impact of CT is
statistically significant and substantial (14\%). Thus, we consider the
model in panel (d) as the most explainable according to Def.
\ref{def-explainability}. Next, we present the results for all remaining
datasets.

@@

\renewcommand{\bibsection}{}
\bibliography{../quarto_ecml/bibliography.bib}
\end{document}

>>>>

Today's prominence of artificial intelligence (AI) has largely been
driven by the success of representation learning with large degrees of
freedom: instead of relying on features and rules hand-crafted by
humans, modern machine learning (ML) models are tasked with learning
highly complex representations directly from the data, guided by narrow
objectives such as predictive accuracy \citep{goodfellow2016deep}. These
models tend to be so complex and opaque that humans cannot easily
interpret their decision-making.

Counterfactual explanations (CE) have become an integral part of the
broader explainable AI (XAI) toolkit \citep{molnar2022interpretable}
that can be employed to make sense of this complexity. Originally
proposed by \citet{wachter2017counterfactual}, CEs prescribe minimal
changes for factual inputs that, if implemented, would prompt some
fitted model to produce an alternative, more desirable output
(Figure~\ref{fig-poc}). This is useful and necessary to not only
understand how opaque models make predictions, but also to provide
recourse to individuals subjected to them: a retail bank, for example,
could use CE to provide meaningful feedback to unsuccessful loan
applicants that were rejected based on an opaque AI decision-making
system (ADM).

For feedback to be meaningful, counterfactual explanations need to
fulfill certain desiderata
\citep{verma2020counterfactual, karimi2020survey}: they should be
faithful to the model \citep{altmeyer2024faithful}, plausible
\citep{joshi2019realistic} and actionable \citep{ustun2019actionable}.
Plausibility is typically understood as counterfactuals being
\emph{in-domain}: unsuccessful loan applicants that implement recourse
should end up with a credit profile that is genuinely similar to that of
individuals who have successfully repaid their loans in the past.
Actionable explanations comply with practical constraints: a young,
unsuccessful loan applicant cannot increase their age in an instance.

Existing state-of-the-art (SOTA) approaches in the field have largely
focused on designing model-agnostic CE methods that identify subsets of
counterfactuals, which comply with specific desiderata. This is
problematic, because a narrow focus on any specific desideratum can
adversely affect others: it is possible, for example, to generate
plausible counterfactuals for models that are also highly vulnerable to
implausible, possibly adversarial counterfactuals
\citep{altmeyer2024faithful}. In this work, we therefore follow the
paradigm that models---as opposed to explanation methods---should be
held accountable for plausible and actionable explanations. While
previous work has shown that at least plausibility can be indirectly
achieved through existing techniques aimed at models' generative
capacity, generalization and robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
we directly incorporate both plausibility and actionability in the
training objective of models to improve their overall explanatory
capacity.

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcome of CT compared to conventionally trained models:
in panel (a), faithful and valid counterfactuals end up near the
decision boundary forming a clearly distinguishable cluster in the
target class (orange); in panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals; in panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features; by contrast,
the decision boundary in panel (c) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful CE with
ground-truth data while protecting immutable features, and capitalize on
methodological links between CE and AE by penalizing the model's
adversarial loss on interim (\emph{nascent}) counterfactuals. To the
best of our knowledge, CT represents the first venture in this direction
with promising results, both empirical and theoretical.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) implausibility of CE decreases by 16\% on average; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 15\%; and (3) robust accuracy increases for all datasets by up to 42
percentage points. Finally, in Section~\ref{sec-discussion}, we discuss
open challenges and conclude.

\caption{\label{fig-poc}Counterfactual explanations (stars) for linear
classifiers trained under different regimes on synthetic data: (a)
conventional training, all mutable; (b) CT, all mutable; (c)
conventional, \emph{age} immutable; (d) CT, \emph{age} immutable. The
linear decision boundary is shown in green along with training data
colored according to ground-truth labels: \(y^-=1\) (blue) and \(y^+=2\)
(orange). Stars indicate counterfactuals in the target class. Synthetic
features are annotated as \emph{debt} and \emph{age} for illustrative
purposes.}

@@

To make the desiderata for our framework more concrete, we follow
previous work in tying the explanatory capacity of models to the quality
of CEs that can be generated for them
\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
we refer to `explanatory capacity' simply as `explainability' in what
follows (see Def. \ref{def-explainability} below for a more formal
exposition).

\subsection{Explainability and Contrastive
Learning}\label{explainability-and-contrastive-learning}

In a closely related work, \citet{altmeyer2024faithful} show that model
averaging and, in particular, contrastive model objectives can produce
more explainable and hence trustworthy models. To do so, the authors
first propose a way to generate counterfactuals that are maximally
faithful to the model in that they are consistent with what the model
has learned about the underlying data. Formally, they rely on tools from
energy-based modelling \citep{teh2003energy} to minimize the divergence
between the distribution of counterfactuals and the conditional
posterior over inputs learned by the model. Their counterfactual
explainer, \emph{ECCCo}, yields plausible explanations if and only if
the underlying model has learned representations that align with them.
The authors find that both deep ensembles
\citep{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\citep{grathwohl2020your}---a form of constrastive learning---tend to do
well in this regard.

It helps to look at these findings through the lens of representation
learning with high degrees of freedom. Deep ensembles are approximate
Bayesian model averages, which are particularly effective when models
are underspecified by the available data \citep{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of
overrelying on a single locally optimal representation that corresponds
to semantically meaningless explanations for the data. Likewise,
previous work of \citet{schut2021generating} found that generating
plausible (``interpretable'') CEs is almost trivial for deep ensembles
that have undergone adversarial training. The case for JEMs is even
clearer: they optimize a hybrid objective that induces both high
predictive performance and strong generative capacity
\citep{grathwohl2020your}, which bears resemblance to the idea of
aligning models with plausible explanations and has inspired our CT
objective.

\subsection{Explainability and Robust
Learning}\label{explainability-and-robust-learning}

\citet{augustin2020adversarial} show that CEs tend to be more meaningful
(``explainable'') if the underlying model is more robust to adversarial
examples. Once again, we can make intuitive sense of this finding if we
look at adversarial training (AT) through the lens of representation
learning with high degrees of freedom: highly complex and flexible
models may learn representations that make them sensitive to implausible
or even adversarial examples \citep{szegedy2013intriguing}. Thus, by
inducing models to ``unlearn'' susceptibility to such examples,
adversarial training can effectively remove implausible explanations
from the solution space.

This interpretation of the link between explainability through
counterfactuals on the one side, and robustness to adversarial examples
on the other is backed by empirical evidence.
\citet{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\citet{abbasnejad2020counterfactual} argue that counterfactuals
represent potentially useful training data in machine learning,
especially in supervised settings where inputs may be reasonably mapped
to multiple outputs. They, too, show that augmenting the training data
of (image) classifiers can improve generalization performance. Finally,
\citet{teney2020learning} argue that counterfactual pairs tend to exist
in training data. Hence, their approach aims to identify similar input
samples with different annotations and ensure that the gradient of the
classifier aligns with the vector between such pairs of counterfactual
inputs using a cosine distance loss function.

@@

@@

@@

\renewcommand{\bibsection}{}
\bibliography{../quarto_aaai/bibliography.bib}
\end{document}
